\propertitle

### 2019-06-25 maxNt_v5.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-06-19T19:49:04+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

In the present work we give an answer to this question: What inferences
about the \emph{total activity} of a large neuronal network can we make by
recording a very small sample thereof? We'll first give an answer by
combining, in a straightforward way, the maximum-entropy method and basic
sampling relations of the probability calculus. Later on we'll show that
this answer can be obtained by exclusively using the probability calculus;
this point of view shows that the maximum-entropy answer is a first
approximation to the answers of \emph{four} different questions, and also
shows the assumptions lying behind it.

We apply our approach to two concrete data sets: the activity of 65 neurons
recorded from a rat's Medial Entorhinal Cortex \mynote{ref}, and the
activity of 159 neurons recorded from a macaque's Motor Cortex
\mynote{ref}\mynote{add recording length}. In the first data set, the most
interesting inference is that the most probable frequency distribution for
the total activity of the full network, if the latter is assumed to be
larger than 1\,000 neurons, has two very distinct modes, both at low
activities ($0.7\%$ and $2.6\%$ of simultaneously active neurons within
$3\;\mathrm{ms}$, see \fig***). The analogous distribution for second data
set doesn't have two modes but still presents one prominent shoulder in its
one low-activity mode. Note that these inferred features for the full
network aren't observed in the sample, nor are they inferred by the
application of maximum-entropy to the sample alone. \mynote{if we manage
  the Monte Carlo sampling or other approx: say something about the
  deviation from the most probable frequency distribution}

@@

The maximum-entropy method \citep{jaynes1957}[much clearer
in][]{jaynes1963}{sivia1996_r2006,hobsonetal1973,jaynes1985b,grandy1980}
has been used for a variety of inferences about the neuronal activity of
various brain areas and about other phenomena of importance to the
neurosciences, for example gene and protein interaction \citep[for
example][]{martignonetal1995,bohteetal2000,shlensetal2006,schneidmanetal2006,tkaciketal2006,mackeetal2009b,tkaciketal2009,roudietal2009c,barreiroetal2010,gerwinnetal2010,mackeetal2011b,ganmoretal2011,cohenetal2011,granotatedgietal2013,mackeetal2013,tkaciketal2014b,shimazakietal2015,moraetal2015,lezonetal2006,weigtetal2009}.
This method is often used to test whether some statistics of the data, for
example second-order time correlations, is sufficient for making other
inferences about the data. It can be considered as an approximation of
probabilistic models based on various assumptions of inferential
sufficiency \citep{jaynes1986d_r1996,portamana2017} -- for example of
empirical moments or of long-run frequencies.

@@

Let's introduce some context and notation for our inference problem.

@@

\mynote{Move this to intro?}We want to stress the usefulness of making
inferences about the full-network activity. First of all, this seems to be
the primary idea behind recording a sample. Second, it allows us to make
comparisons across experimental sessions; such comparisons would me
difficult or meaningless made at the level of recorded samples, which
generally comprise non-overlapping sets of neurons differing also in size.

@@

Regarding~\ref{item:maxent_step}, we assume that you're familiar with the
maximum-entropy method. In short, it amounts to a pair of prescriptions:
choose the distribution having maximum Shannon entropy among those
satisfying specific convex constraints, such as fixed averages; and judge
these averages to be equal to measured averages. We add two remarks about
this method that are seldom made in the literature. First, the distribution
$\px(\yA)$ given by this method is the zeroth-order approximation \citep[in
the sense of Laplace's
method:][\chap~4]{debruijn1958_r1961}{tierneyetal1986,strawderman2000} of
four different distributions for the full network:
\begin{itemize}
\item the most probable \emph{frequency} distribution for the total
  activity across the \emph{recorded} bins, %$\yFF(\yA)$,
\item the most probable \emph{frequency} distribution for the total
  activity in a very long run of \emph{new} time bins, %$\ynuu(\yA)$,
\item the \emph{probability} distribution for the value of the total
  activity at any time bin among those \emph{recorded},
  %$\yp(\yA_{t} \| \yH)$ with $1\le t \le T$,
\item the \emph{probability} distribution for the value of the total
  activity at a \emph{new} time bin.
  % $\yp(\yA_{t} \| \yH)$ with $t<1$ or $t>T$
\end{itemize}
The maximum-entropy distribution is thus an approximate answer to four
completely different inferences. Note that the four distributions above
numerically differ in higher-order approximations. We discuss their
differences in \sect***. Second, the maximum-entropy method based on the
Shannon entropy implicitly makes some assumptions about the probabilities
for the long-run frequency distributions
\citep{jaynes1986d_r1996,portamana2009,portamana2017}. We discuss these
assumptions in \sect***.

@@

The number of moments used with this method depends on the questions and
hypotheses that a researcher is exploring; for example, hypotheses of
sufficient statistics, such as the sufficiency of pairwise correlations for
inferential purposes. In the present work we only want to present the
general method without entering into biological questions of this kind.

The particular case in which $n$ moments are constrained is especially
important: it corresponds to fully constraining the marginal frequency
distribution for the activity of the sample neurons. In this case, our
inferences about the full-network activity are using all available measured
frequency data. Note that application of the maximum-entropy method
\emph{at the sample level} is trivial and meaningless in this case -- it
just gives back the measured frequency distribution. But application of the
method \emph{at the level of the full network} is not trivial.

@@

We first calculate the distribution by using six moments. This number
already provides as much information as the full frequency distribution of
the sample, and at the same time illustrates the use of the approach in
questions of statistic sufficiency (typically limited to two or three
moments). Figure~\ref{fig:plots_6mom} shows the resulting densities
(distribution/activity-step, that is, distribution${}\times N$) for
full-network sizes $N=n$, $N=1\,000$, $N=5\,000$,
$N=10\,000$\mynote{motivate?}. The case $N=n$ corresponds to applying the
maximum-entropy method at the sample level; it can be observed that with
six moments it reproduces almost exactly the measured frequency
distribution.
\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{scripts/paper_stensola_6mom_distributionsN_prior2.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{scripts/paper_riehle_6mom_distributionsN_prior2.pdf}%
\caption{***}
\label{fig:plots_6mom}
\end{figure}
%maxent_solutions_different_moments_prior2_6moments.nb

@@

To illustrate how our approach can be applied to studies of sufficient
statistics, \fig~\ref{fig:plots_2_4mom} shows the results for two
constrained moments (equivalent to constraining means and correlations
only) and for four constrained moments, with $N=10\,000$. In either data
set we obtain two very distinct distributions. For the first data set in
particular, four moments lead to a bimodal distribution, whereas two to a
unimodal one, showing that two moments would not be a sufficient
statistics. For comparison, the two distributions obtained applying
maximum-entropy \emph{at the sample level} are shown as an inset in the
plot for the first data set: their differences aren't so glaring as in the
full-network application.
\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{scripts/paper2_stensola_twomom_distributions10000_prior2.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{scripts/paper_riehle_twomom_distributions10000_prior2.pdf}%
\caption{***}
\label{fig:plots_2_4mom}
\end{figure}
%maxent_solutions_different_moments_prior2_6moments.nb

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-06-25T10:55:38+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

In the present work we give an answer to this question: How much is our
uncertainty about the \emph{total activity} of a large neuronal network, if
we have observed the activity of a very small sample thereof? We'll first
give an answer by combining, in a straightforward way, the maximum-entropy
method and basic sampling relations of the probability calculus. Later on
we'll show that this answer can be obtained by exclusively using the
probability calculus; this point of view shows that the maximum-entropy
answer is a first approximation to the answers of \emph{four} different
questions, and also shows the assumptions lying behind it.

We apply our approach to two concrete data sets: the activity of 65 neurons
recorded from a rat's Medial Entorhinal Cortex \mynote{ref}, and the
activity of 159 neurons recorded from a macaque's Motor Cortex
\mynote{ref}\mynote{add recording length}. In the first data set, the most
interesting finding is that the most probable frequency distribution for
the total activity of the full network, if the latter is assumed to be
larger than 1\,000 neurons, has two very distinct modes, both at low
activities ($0.7\%$ and $2.6\%$ of simultaneously active neurons within
$3\;\mathrm{ms}$, see \fig***). The analogous distribution for second data
set doesn't have two modes but still presents one prominent shoulder in its
one low-activity mode. Note that these inferred features for the full
network aren't observed in the sample, nor are they inferred by the
application of maximum-entropy to the sample alone. \mynote{if we manage
  the Monte Carlo sampling or other approx: say something about the
  deviation from the most probable frequency distribution}

@@

The maximum-entropy method \citep{jaynes1957}[much clearer
in][]{jaynes1963}{sivia1996_r2006,hobsonetal1973,jaynes1985b,grandy1980}
has been used for a variety of forecasts about the neuronal activity of
various brain areas and about other phenomena of importance to the
neurosciences, for example gene and protein interaction \citep[for
example][]{martignonetal1995,bohteetal2000,shlensetal2006,schneidmanetal2006,tkaciketal2006,mackeetal2009b,tkaciketal2009,roudietal2009c,barreiroetal2010,gerwinnetal2010,mackeetal2011b,ganmoretal2011,cohenetal2011,granotatedgietal2013,mackeetal2013,tkaciketal2014b,shimazakietal2015,moraetal2015,lezonetal2006,weigtetal2009}.
This method is often used to test whether some statistics of the data, for
example second-order time correlations, is sufficient for quantifying our
uncertainty about the data. It can be considered as an approximation of
probabilistic models based on various assumptions of inferential
sufficiency \citep{jaynes1986d_r1996,portamana2017} -- for example of
empirical moments or of long-run frequencies.

@@

Let's introduce some context and notation for our problem.

@@

\mynote{Move this to intro?}We want to stress the usefulness of making
quantified guesses about the full-network activity. First of all, this seems to be
the primary idea behind recording a sample. Second, it allows us to make
comparisons across experimental sessions; such comparisons would me
difficult or meaningless made at the level of recorded samples, which
generally comprise non-overlapping sets of neurons differing also in size.

@@

Regarding~\ref{item:maxent_step}, we assume that you're familiar with the
maximum-entropy method. In short, it amounts to a pair of prescriptions:
choose the distribution having maximum Shannon entropy among those
satisfying specific convex constraints, such as fixed averages; and judge
these averages to be equal to measured averages. We add two remarks about
this method that are seldom made in the literature. First, the distribution
$\px(\yA)$ given by this method is the zeroth-order approximation \citep[in
the sense of Laplace's
method:][\chap~4]{debruijn1958_r1961}{tierneyetal1986,strawderman2000} of
four different distributions for the full network:
\begin{itemize}
\item the most probable \emph{frequency} distribution for the total
  activity across the \emph{recorded} bins, %$\yFF(\yA)$,
\item the most probable \emph{frequency} distribution for the total
  activity in a very long run of \emph{new} time bins, %$\ynuu(\yA)$,
\item the \emph{probability} distribution for the value of the total
  activity at any time bin among those \emph{recorded},
  %$\yp(\yA_{t} \| \yH)$ with $1\le t \le T$,
\item the \emph{probability} distribution for the value of the total
  activity at a \emph{new} time bin.
  % $\yp(\yA_{t} \| \yH)$ with $t<1$ or $t>T$
\end{itemize}
The maximum-entropy distribution is thus an approximation of our
uncertainty about four completely different quantities. Note that the four
distributions above numerically differ in higher-order approximations. We
discuss their differences in \sect***. Second, the maximum-entropy method
based on the Shannon entropy implicitly makes some assumptions about the
probabilities for the long-run frequency distributions
\citep{jaynes1986d_r1996,portamana2009,portamana2017}. We discuss these
assumptions in \sect***.

@@

The number of moments used with this method depends on the questions and
hypotheses that a researcher is exploring; for example, hypotheses of
sufficient statistics, such as the sufficiency of pairwise correlations to
quantify our uncertainty. In the present work we only want to present the
general method without entering into biological questions of this kind.

The particular case in which $n$ moments are constrained is especially
important: it corresponds to fully constraining the marginal frequency
distribution for the activity of the sample neurons. In this case, our
uncertaity about the full-network activity is based on all available
measured frequency data. Note that application of the maximum-entropy
method \emph{at the sample level} is trivial and meaningless in this case
-- it just gives back the measured frequency distribution. But application
of the method \emph{at the level of the full network} is not trivial.

@@

We first calculate the distribution by using six moments. This number
already provides as much information as the full frequency distribution of
the sample, and at the same time illustrates the use of the approach in
questions of statistic sufficiency (typically limited to two or three
moments). Figure~\ref{fig:plots_6mom} shows the resulting densities
(distribution/activity-step, that is, distribution${}\times N$) for
full-network sizes $N=n$, $N=1\,000$, $N=5\,000$,
$N=10\,000$\mynote{motivate?}. The case $N=n$ corresponds to applying the
maximum-entropy method at the sample level; it can be observed that with
six moments it reproduces almost exactly the measured frequency
distribution.
\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/paper_stensola_6mom_distributionsN_prior2.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/paper_riehle_6mom_distributionsN_prior2.pdf}%
\caption{***}
\label{fig:plots_6mom}
\end{figure}
%maxent_solutions_different_moments_prior2_6moments.nb

@@

To illustrate how our approach can be applied to studies of sufficient
statistics, \fig~\ref{fig:plots_2_4mom} shows the results for two
constrained moments (equivalent to constraining means and correlations
only) and for four constrained moments, with $N=10\,000$. In either data
set we obtain two very distinct distributions. For the first data set in
particular, four moments lead to a bimodal distribution, whereas two to a
unimodal one, showing that two moments would not be a sufficient
statistics. For comparison, the two distributions obtained applying
maximum-entropy \emph{at the sample level} are shown as an inset in the
plot for the first data set: their differences aren't so glaring as in the
full-network application.
\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/paper2_stensola_twomom_distributions10000_prior2.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/paper_riehle_twomom_distributions10000_prior2.pdf}%
\caption{***}
\label{fig:plots_2_4mom}
\end{figure}
%maxent_solutions_different_moments_prior2_6moments.nb
